---
title: "Patient Treatment Timelines for Longitudinal Survival Data"
author: "Katherine Hoffman"
date: 2019-11-28T21:13:14-05:00
categories: ["R"]
tags: ["R"]
output: 
  html_document:
    toc: true
    toc_float: true
    smart: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="asis")
```

I am a biostatistician at a research university, and I often find myself working with longitudinal survival data. As with any data analysis, I need to examine the quality of my data before deciding which statistical methods to implement. 

This post contains reproducible examples for how I prefer to visually explore survival data containing longitudinal exposures or covariates. I create a "treatment timeline" for each patient, and the end product looks something like this:

```{r, echo=F, warning=F, message=F, fig.height=6.5, fig.width=10}
library(tidyverse, warn.conflicts = F)
set.seed(777)
n <- 50
dat <- 
  tibble(.rows = n) %>%
  mutate(pt_id = factor(row_number()),
         months_followup = round(runif(n, 0, 12)),
         pt_trt_prob = runif(n, 0, 1),
         death = rbinom(n, 1, 0.5)) %>%
  group_by(pt_id) %>%
  complete(months_followup = full_seq(0:max(months_followup), 1)) %>%
  fill(pt_trt_prob, .direction = "up") %>%
  mutate(death = replace_na(death, 0),
         last_month_followup = max(months_followup),
         death_date = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ungroup() %>%
  mutate(trt = rbinom(row_number(), 1, pt_trt_prob)) #%>%
  #mutate(trt  = ifelse(sample_n(., 30)$pt_id == pt_id, NA, trt)) 

dat_cc <- 
  tibble(.rows = n) %>%
  mutate(pt_id = row_number(),
         months_followup = round(runif(n, 0, 12)),
         pt_cov_mean = runif(n, 80, 150),
         death = rbinom(n, 1, 0.5)) %>%
  group_by(pt_id) %>%
  complete(months_followup = full_seq(0:max(months_followup), 1)) %>%
  fill(pt_cov_mean, death, .direction = "up") %>%
  mutate(last_month_followup = max(months_followup),
         death_date = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ungroup() %>%
  mutate(cov = rnorm(row_number(), pt_cov_mean, 10))

dat_cc_miss <-
  dat_cc %>%
    group_by(pt_id, last_month_followup, pt_cov_mean) %>%
    nest() %>% # one row per patient
    mutate(last_month_followup_bound =
             ifelse(last_month_followup > 1, last_month_followup - 1, last_month_followup)) %>%
    ungroup() %>% 
    mutate(n_pt = round(runif(n(), 0, round(last_month_followup_bound - 3*(1/pt_cov_mean))))) %>% # add sample sizes
  mutate(samp = map2(data, n_pt, sample_n)) %>% 
  select(-data) %>%
  unnest() %>%
  select(pt_id, months_followup, cov) %>%
  right_join(dat_cc %>% group_by(pt_id) %>% filter(row_number() == 1) %>% select(pt_id, last_month_followup, death) %>% ungroup()) %>%
  group_by(pt_id) %>%
  complete(months_followup = full_seq(0:max(last_month_followup), 1)) %>% 
  arrange(pt_id, last_month_followup) %>%
  fill(last_month_followup, .direction = "down") %>%
  mutate(last_month_followup = replace_na(last_month_followup, max(months_followup))) %>%
  ungroup()

p <- dat %>%
  mutate(trt = factor(trt, levels=0:1, labels=c("No","Yes"))) %>%
  mutate(pt_id_ordered = fct_rev(fct_reorder(pt_id, last_month_followup)),
         death = fct_drop(factor(death))) %>%
  ggplot(aes(x = months_followup,
             y = pt_id_ordered,
             group = pt_id_ordered,
             col = trt)) +
  geom_line()  + 
  geom_point(shape = 15) + 
  geom_point(aes(x = death_date, y = pt_id_ordered), shape=4, col="black") +
  theme_bw() +
  labs(x = "Months of Follow-Up",
       y = "Patient ID",
       col = "Treatment",
       title = "Patient Timelines: Binary Treatment",
       subtitle = "x indicates time of patient death") +
  theme(axis.text.y = element_text(size=6),
        legend.position = c(.7, .9),
        legend.direction = "horizontal",
        legend.background = element_rect(linetype="solid", colour ="black")) +
  scale_x_continuous(expand=c(0.01,0.01)) +
  scale_color_manual(values=c("dodgerblue","firebrick1")) 
p2 <- dat_cc_miss %>%
  mutate(pt_id = fct_rev(fct_reorder(factor(pt_id), last_month_followup)),
  month_death = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ggplot(aes(x = months_followup,
             y = pt_id,
             group = pt_id,
             col = cov)) +
  geom_line()  + 
  geom_point(shape = 15) + 
  geom_point(aes(x = month_death, y = pt_id), shape=4, col="black") +
  theme_bw() +
  labs(x = "Months of Follow-Up", y = "Patient ID", col = "Covariate",
       title = "Patient Timelines: Continuous Covariate",
       subtitle = "x indicates month of patient death, gray indicates missing covariate") +
  theme(axis.text.y = element_text(size=6),
        legend.position = c(.7, .9),
        legend.direction = "horizontal",
        legend.background = element_rect(linetype="solid", colour ="black")) +
  scale_x_continuous(expand=c(0.01,0.01)) +
  scale_color_gradient(low="dodgerblue",high="firebrick1",na.value = "lightgray") 
gridExtra::grid.arrange(p,p2,nrow=1)
```

Each line represents one patient, and each square represents a measurement at a specified point. I find these graphs useful to look for patterns or discrepancies in follow up times, treatment patterns, and missingness. They also allow me to verify that my coding is correct as I move through different data manipulations for my analyses.

For the following examples I generate longitudinal survival data sets. I skip over how I made this data because it's not the focus of this post, but if you have questions about the data generation please let me know. In general, I am trying to work from the point in a survival data analysis where you've already calculated each individual's time to event.

# Example 1: Binary Treatment

In this first example, we'll be generating months until death or censoring, which are dependent on a binary treatment.

To make these charts, my first step is to load the package `tidyverse`, since we'll need functions from `dplyr`, `tidyr`, `forcats`, and `ggplot2`. I then set a seed for data generation. Finally, I set my `n` to the number of patients I want in my data. The data sets I make these visualizations for are typically 100-500 patients in size, but for space purposes I'll set my `n` to only 50.

```{r}
library(tidyverse)
set.seed(7)
n <- 50
```

Here is the data generation code. If you're interested in how I made it, you can check the comments, but if not, **I recommend just copying and pasting into your `R` console to try the plotting code!**

```{r}
dat <- 
  tibble(.rows = n) %>%  # empty data frame / tibble with n rows
  mutate(pt_id = factor(row_number()),  # patient ids are 1-n
         pt_trt_prob = runif(n,0,1), # randomly generate a treatment probability for each patient
         months_followup = round(runif(n, 0, 20) + 3 * pt_trt_prob), # randomly generate length of time in the study, which is affected by the prob of treatment
         death = rbinom(n, 1, 1-pt_trt_prob)) %>% # death is affected by prob of treatment
  group_by(pt_id) %>%   # group by patient so we can do more data manipulation
  complete(months_followup = full_seq(0:max(months_followup), 1)) %>% # add in all the months patients are in the study
  fill(pt_trt_prob, .direction = "up") %>% # fill in the treatment probability I made earlier so I can use this to add treatment for every time point
  ungroup() %>% # no longer need patients grouped
  mutate(trt = factor(rbinom(row_number(), 1, pt_trt_prob)),
         death = replace_na(death, 0)) %>%   # fill in treatment for everyone based on their treatment probability
  select(pt_id, trt, months_followup, death) # remove leftover columns from data generation

```

Let's look at the data. It's in "long" format for each month each subject had a measurement from.

```{r}
knitr::kable(head(dat))
```

We can now plot our data in a very basic way. We really only need to specify in the `aes`thetics that the `x`-axis is time, the `y`-axis is subject IDs, the `col`or should correspond to treatment, and *our lines should be `group`ed together by subject*. Don't forget that last one! Then we can say we want `geom_line` to make a base timeline for each subject, and the points on the timeline should be squares (`shape = 15` in the `geom_point` mapping function).

Et voilà! A cute patient treatment timeline.

```{r}
dat %>%
  ggplot(aes(x = months_followup, y = pt_id, group = pt_id, col = trt)) +
  geom_line()  + 
  geom_point(shape = 15)
```

Okay, it's not *that* cute. But that little bit of code is really the core of the plot! If you want to add markers of death (or any other event) and rearrange by length of follow up time it just takes a little bit of extra data manipulation and one additional line of `ggplot2` code.

```{r}
dat %>%
  group_by(pt_id) %>%
  # make a new column with all patients last follow up visit
  mutate(last_month_followup = max(months_followup), 
         # new variable for month that patients died, if they died
         month_death = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
         # reorder pt id by last month of follow up (highest to lowest)
         # without fct_rev, chart is arranged in opposite direction (lowest to highest)
  ungroup() %>%
  mutate(pt_id = fct_rev(fct_reorder(pt_id, last_month_followup))) %>%
  ggplot(aes(x = months_followup, y = pt_id, group = pt_id, col = trt)) +
  geom_line()  + 
  geom_point(shape = 15) + 
  # add in a new mapping layer of points that correspond to death
  geom_point(aes(x = month_death, y = pt_id), col = "black", shape = 4) 
```

Don't worry about the warning for removing missing values -- it's because we have `NA`s at most months for our `month_death`, so `geom_point` doesn't know where to put an 'x' marker.

Finally, let's work on making the plot a bit nicer to look at:

```{r}
dat %>%
  group_by(pt_id) %>%
  mutate(last_month_followup = max(months_followup), 
         month_death = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ungroup() %>%
  mutate(pt_id = fct_rev(fct_reorder(pt_id, last_month_followup)),
         # make the treatment variable labels nicer
         trt = factor(trt, levels=0:1, labels=c("No","Yes"))) %>%
  ggplot(aes(x = months_followup, y = pt_id, group = pt_id, col = trt)) +
  geom_line()  + 
  geom_point(shape = 15) + 
  geom_point(aes(x = month_death, y = pt_id), col = "black", shape = 4) +
  theme_bw() +
  labs(x = "Months of Follow-Up", y = "Patient ID", col = "Treatment",
       title = "Patient Treatment Timeline",
       subtitle = "x indicates time of patient death") +
  # edit legend box and make patient ids small
  theme(axis.text.y = element_text(size=6),
        legend.position = c(.6, .9), legend.direction = "horizontal",
        legend.background = element_rect(linetype="solid", colour ="black")) +
  # remove extra space around timeline
  scale_x_continuous(expand=c(0.01,0.01)) +
  # set the color of the lines and points
  scale_color_manual(values=c("dodgerblue","firebrick1"))

```

# Example 2: Continuous Covariate with Missingness

We can follow the very same steps for making a timeline for a continuous variable.

Again, here is the data generation code. 

```{r, eval=F}
dat_cc <- 
  tibble(.rows = n) %>%
  mutate(pt_id = row_number(),
         months_followup = round(runif(n, 0, 12)),
         pt_cov_mean = runif(n, 80, 150),
         death = rbinom(n, 1, 0.5)) %>%
  group_by(pt_id) %>%
  complete(months_followup = full_seq(0:max(months_followup), 1)) %>%
  fill(pt_cov_mean, death, .direction = "up") %>%
  mutate(last_month_followup = max(months_followup),
         death_date = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ungroup() %>%
  mutate(cov = rnorm(row_number(), pt_cov_mean, 10)) 
```

This time I am incorporating missingness at a patient-level on my simulated covariate measurements to show how we can use the plots to look at potential patterns in missingness.

```{r}
dat_cc_samp <- dat_cc %>%
  mutate(idx = row_number()) %>%
  sample_frac(.5, weight = pt_cov_mean^3)  %>% # sample 50% of data, with weights for the sample determined by the patient's mean covariate. This would mean patients with a higher covariate are more likely to have missing data. Perhaps they are sicker patients!
  pull(idx)

dat_cc_miss  <- dat_cc %>%
  mutate(cov = case_when(row_number() %in% dat_cc_samp ~ NA_real_,
                         TRUE ~ cov)) %>%
  select(pt_id, months_followup, cov, death, last_month_followup)
```

Looking again at the data we'll be using:

```{r}
knitr::kable(head(dat_cc_miss))
```

We can plot the data and see this missingness with gray timeline points and lines:

```{r}
dat_cc_miss %>%
  mutate(pt_id = fct_rev(fct_reorder(factor(pt_id), last_month_followup)),
  month_death = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ggplot(aes(x = months_followup, y = pt_id,
             group = pt_id, col = cov)) +
  geom_line()  + 
  geom_point(shape = 15) + 
  geom_point(aes(x = month_death, y = pt_id), shape=4, col="black") +
  theme_bw() +
  labs(x = "Months of Follow-Up", y = "Patient ID", col = "Covariate",
       title = "Patient Timelines: Continuous Covariate",
       subtitle = "x indicates month of patient death, gray indicates missing covariate") +
  theme(axis.text.y = element_text(size=6),
        legend.position = c(.7, .9),
        legend.direction = "horizontal",
        legend.background = element_rect(linetype="solid", colour ="black")) +
  scale_x_continuous(expand=c(0.01,0.01)) +
  scale_color_gradient(low="dodgerblue",high="firebrick1",na.value = "lightgray") 
```

Alternatively, we could rearrange our timelines by the number of measures we have for each patients `cov`ariate of interest by making a new column containing the `sum(!is.na(cov))` and rearranging our `pt_id` by that column.

```{r, fig.width=8}
dat_cc_miss %>%
  group_by(pt_id) %>%
  mutate(n_measures = sum(!is.na(cov))) %>%
  ungroup() %>%
  mutate(
    pt_id = fct_rev(fct_reorder(factor(pt_id), n_measures)),
  month_death = case_when(death == 1 ~ last_month_followup, TRUE ~ NA_real_)) %>%
  ggplot(aes(x = months_followup, y = pt_id,
             group = pt_id, col = cov)) +
  geom_line()  + 
  geom_point(shape = 15) + 
  geom_point(aes(x = month_death, y = pt_id), shape=4, col="black") +
  theme_bw() +
  labs(x = "Months of Follow-Up", y = "Patient ID", col = "Covariate",
       title = "Patient Timelines: Continuous Covariate",
       subtitle = "x indicates month of patient death, gray indicates missing covariate") +
  theme(axis.text.y = element_text(size=6),
        legend.position = c(.7, .9),
        legend.direction = "horizontal",
        legend.background = element_rect(linetype="solid", colour ="black")) +
  scale_x_continuous(expand=c(0.01,0.01)) +
  scale_color_gradient(low="dodgerblue",high="firebrick1",na.value = "lightgray") 
```

From this plot we can start to see the data generating mechanism that patients with less missing measurements measurements tend to have longer lengths of follow up and lower covariate measurements. If this were my real data, I would follow up with my research collaborators for more information on this missingness pattern.

Please let me know if you have questions or a suggestion for a data set. Happy treatment timeline plotting!

